---
title: "R Notebook"
output: html_notebook
---

# Summary

Hierarchical or multilevel modeling is a generalization of regression modeling.

Rationale of multilevel modeling comes from the dependencies of observations within groups.

In cases where there are well-defined hierarchical structures in the data, there are also multilevel dependencies between samples, and hierarchical or mixed effects model should be preferred.
If we opt to pool all samples together and fit a fixed effect model in such cases, we ignore the group information, and won't be able to make group-specific predictions.
If we opt to estimate group-specific effects by fitting a fixed effect model, we are risky of overfitting in groups of few samples, and consequences poor out-of-sample predictions. 

Additional assumption of the random term is the modeling of the group-level effects, such as to fit the group-level effects using a Gaussian model, which might not hold.

In random effect model, we still estimate group-specific parameters, but these parameters are further modeled by hyperparameters.
As a result, the group-specific parameter estimand are results of two sources: 1, data within the group; 2, data of other groups also play a role. 

# Case study using radon contamination data

```{r}
library(rstan)
load("~/GitHub/stats/mixed/radon.rdt")
attach(radon)

hist(log.radon)
table(basement)

unique(uranium[county == 1])
unique(uranium[county == 2])
```

### complete pool

```{r}
pool = stan_model("~/GitHub/stats/mixed/pool.stan")

x <- cbind(1, radon$basement)
data <- list(N = dim(x)[1], D = dim(x)[2], x = x, y = radon$log.radon)

fit.pool <- sampling(pool, data = data)
w <- extract(fit.pool)$w

plot(basement, log.radon); lines(basement, x %*% apply(w, 2, mean), col='red')
```

### mixed effect model (partial pooling)

```{r}
mixed = stan_model("~/GitHub/stats/mixed/mixed.stan")

data$C = length(unique(county))
data$l = county
fit.mixed <- sampling(mixed, data = data)
w <- extract(fit.pool)$w

```

> how to use HLM results to make predictions on a new group
> likely expected mean of the group-specific parameters

> what is the expected mean of the group-specific parameters?
> it deviates a bit, and is always between the overall expected mean and the within group expected mean. 

```{r}
print(fit.mixed, pars = "mu_w")
print(fit.mixed, pars = c("w[2,1]", "w[2,2]"))

radon.1 =  radon[radon$county == 2, ]
summary(lm(log.radon ~ basement, data = radon.1))

```
