# Deep dive why mixed effect models work in estimating heritability

## Mixed effect models are equivalent to Bayesian linear model 

## simulate genotype and phenotypes 

1. Make 5 variants in 3 LD blocks and 2 causal variants
2. Phenotype follows: $$y = \sum_{c \in C}\beta_c*g_c + N(\epsilon|0, \sigma^2I)$$

```{r}
g1 = rbinom(1e3, 2, .5)
noise = which(rbinom(1e3, 1, .1) == 1)
g2 = g1; g2[noise] = 2 - g2[noise]

g3 = rbinom(1e3, 2, .2)
noise = which(rbinom(1e3, 1, .1) == 1)
g4 = g3; g4[noise] = 2 - g4[noise]

g5 = rbinom(1e3, 2, .3)

g = cbind(g1, g2, g3, g4, g5)
library(lattice); levelplot(cor(g))

gc = g[, c(2, 4)] # causal variants
fc = colMeans(gc) / 2
bc = c(1, 2) # effect sizes

e = rnorm(1e3, 0, 2)
var(e)

y = gc %*% bc + e

var(y)

2 * fc * (1-fc) # this is an approximation
apply(gc, 2, var)

sum(2 * fc * (1 - fc) * bc^2)
sum(2 * fc * (1 - fc) * bc^2) + var(e)
sum(apply(gc, 2, var) * bc^2)
sum(apply(gc, 2, var) * bc^2) + var(e)

summary(lm(y ~ g))
summary(lm(y ~ g))$sigma

(h = 1 - var(e) / var(y))
```

## Mixed effect model (Bayesian model) approach

```{r}
K = g %*% t(g) / 5

library(rstan)
model = stan_model("~/Git/mymsd/lmm.stan")

data = list(N = 1e3, pheno = c(y), mymean = rep(0, 1e3), K = K)
myfit = sampling(model, data = data, chains = 2, iter = 2e2) 
```

## K above was oversimplified and problematic, need more real geneomic data to make it work
## it will be interesting to study the variance decomposition schemes in hierarchical modeling framework using examples in BDA

# Models

```{r}
stan_multi_norm <- "

data {
int<lower=1> K; # outcomes
vector[K] mu;
cov_matrix[K] Sigma;
vector[K] y;
}

parameters {
vector[K] beta; 
}

model {
beta ~ multi_normal(mu, Sigma);
}

"

stan_multi_cholesky <- "

data {
int<lower=1> K; # outcomes
vector[K] mu;
cov_matrix[K] Sigma;
vector[K] y;
}

transformed data {
matrix[K, K] L;
L <- cholesky_decompose(Sigma);
}

parameters {
vector[K] beta; 
}
model {
beta ~ multi_normal_cholesky(mu, L);
}
"

load("adsp.rdt")
mdata <- adsp$mdata
Sigma <- adsp$kinship$autosome
Sigma[Sigma < 0] <- 0

stan_multi_norm <- stan_model(model_code = stan_multi_norm)
stan_multi_cholesky <- stan_model(model_code = stan_multi_cholesky) 

dat <- list(K = 576, mu = rep(0, 576), y = mdata$AD1, Sigma = Sigma) 

fit_multi_norm <- sampling(stan_multi_norm, chain = 2, data = dat) # 450 sec each chain
fit_multi_cholesky <- sampling(stan_multi_cholesky, data = dat) # 170 sec each chain 
```