---
title: "ABF"
output: html_document
---

## Wakefield ABF (approximate bayes factor)?

>> Jon Wakefield 2008. Genetic Epidemiology
>> Bayes factors for genome-wide association studies: comparison with P-values

## P-values in genetic association studies

1. A arbituary P-value cutoff to choose significant findings does not reflect statistical power of the given test, which is a function of maf and sample size. 
2. Therefore, this practice punish studies of smaller sample size, likewise the rare variants.

## Bayes factor

$$BF = \frac{Pr(y|H_0)}{Pr(y|H_1)}$$

$$\frac{Pr(H_0|y)}{Pr(H_1|y)} = \frac{Pr(y|H_0)}{Pr(y|H_1)} * \frac{Pr(H_0)}{Pr(H_1)}$$

Significance measure: $\frac{Pr(H_0|y)}{Pr(H_1|y)}$

## Wakefield Approximate Bayes Factor to against the null hypothesis

$$WAFB = \sqrt{V / (V + W)} e^{(Wz^2 / 2 (V+W))}$$

## ?s

>> Why it is better than P-values?
>> How to understand ABF depends on both p-value and power?
>> How to understand when power is high, probability of null and alternative hypothesis are both high?

P(y | H0) and P(y | H1) are technically marginal likelihoods of H0 model and H1 model.

## Wald statistics

```{r}
x = rnorm(1e2, 0, 1)
y = x + rnorm(1e2, 0, 2)

(fit = summary(lm(y ~ x)))
coef = fit$coefficients
wald = coef["x", "Estimate"] / coef["x", "Std. Error"]
wald = wald^2
coef["x", "t value"] ^ 2
1 - pchisq(wald, df = 1)
```

## Bayesian finemapping of association signals: use ABF to identify 95% credible set in finemapping

This script was from Corneliu.

## Get the Credible set for one locus with single signal

```{r}
fm <- read.table("/work/genetics/projects/Gen187_GeneticsDB/GWAS_Summary_stats/Cdiff/data/rs2516513_1Mb_SummaryStatistics.txt", header = T, fill = T)
```

## Calculate approximate bayes factor from effect size (beta) and standard error 
## W is prior on variance of effect size
## Method from Wakefield Gen Epi 2009
## Suggested value for W is 0.04

```{r}
abf <- function(beta,se,W){
  V = se**2
  r = W/(V+W)
  z2 = (beta/se)**2
  abf = ((1-r)**0.5) * exp((z2/2)*r)
  return(abf)
}

fm$abf <- abf(beta=fm$GT_beta, se=fm$GT_beta_se, W=0.04)
```

## Assumes one causal variant and equal priors on all variants
## From Maller et al Nat Gen 2012

```{r}
fm$post = fm$abf/sum(fm$abf)
```

## Identify 95% credible set

```{r}
fm <- fm[order(fm$post, decreasing = T),]
fm$cumpost <- cumsum(fm$post)
credset <- fm[1:min(which(fm$cumpost>0.95)),]
```
