---
title: "The GNetLMM method for trans-eQTL analysis"
output: html_document
---

Rakitsch and Stegle 2016. Genome Biology

```{r}
a = rnorm(1e3, 0, 1)
b = rnorm(1e3, 0, 1)
c = a + b + rnorm(1e3, 0, 1)

summary(lm(c ~ a + b)) # true model

summary(lm(c ~ a)) # correlated a and c
summary(lm(a ~ c)) # correlated a and c

summary(lm(a ~ b))
summary(lm(a ~ b + c))
```

## A and B become correlated conditional on the shared effector C
## Linear regression generates funny results with wrong models
## In this case, B and C are correlated but were treated as independent variables to predict A

```{r}
plot(make_graph(c("A", "C", "B", "C", "C", "D")))

plot(b, a)
f = summary(lm(a ~ b + c))

ac = c * f$coefficients["c", "Estimate"]
plot(b, c)
plot(b, ac)

summary(lm((a - ac) ~ b))
plot(b, (a - ac))
```

## B and D have different correlation relationships with A and C

```{r}
d = c + rnorm(1e3, 0, 1)
summary(lm(a ~ d))
summary(lm(b ~ d))
summary(lm(a ~ d + c))
```

## low-rank LMM implementation

$$y_C \sim N(x_A \beta, \sigma_{exo}^2 \sum y_B y_B^T + \sigma_n^2I)$$
## Simulate data

```{r}
n <- 5 # variables
n2 <- 2e2 # sample size

V <- diag(1, n, n)
x <- rmvn(n2, rep(0, n), V)

b <- rnorm(n, 0, 1)
v <- rnorm(n2, 0, 1)
y <- v + x %*% b + rnorm(n2, 0, 1)
```

## Fixed effect model

```{r}
(f = summary(lm(y ~ v + x)))
plot(f$coefficients[-1, "Estimate"], c(1, b)); abline(0, 1)
```

## Low-rank random effect model using known variance-covariance matrix

```{r}
library(rstan)

mod = stan_model("~/Git/mymsd/gnetlmm.stan")
dat = list(N = n2, x = v, y = c(y))

K = x %*% t(x)
L = t(chol(K, pivot = T)) # an issue
dat$L = L

f2 = sampling(mod, dat, chains = 4, cores = 4)

print(f2, par = c("a", "beta", "sigma_e", "sigma_g"))
plot(f2, par = c("a", "beta", "sigma_e", "sigma_g"))

fs.s = as.data.frame(f2)
summary(fs.s$beta)
hist(fs.s$beta)
```

## Not as accurate, but close enough!
## Does the random effect estimates help?

```{r}
K = diag(1, n2)
L = t(chol(K, pivot = T)) # an issue
dat$L = L

f3 = sampling(mod, dat, chains = 4, cores = 4)

print(f2, par = c("a", "beta", "sigma_e", "sigma_g"))
print(f3, par = c("a", "beta", "sigma_e", "sigma_g"))
```

# Helps
