---
title: "Use GWAS Summary Statistics"
output: html_document
---

## LMM model

A fixed effect polygenic effect model: $y \sim N(\mu + X\beta, \sigma_r^2 I)$, where:
y is a N X 1 vector of phenotype, X is a N X M matrix of causal variants, $\beta$ is a M X 1 vector of the effects of each variant, $\sigma^2$ is the residual variance.

We often do not know all the causal variants, which leads to random effect model. 
Key assumption of random effect model is assuming the elements of $\beta$ follow i.i.d. of $N(0, \sigma_g^2/M)$.

The model has become a two-level hierarchical model. This model is useful in heritability estimation, sample relatedness correction, et al.

## LD matrix

Let V be the M X M LD matrix of the M variants, $V = X^TX / (N-1)$

```{r}
x = rbinom(1e3, 2, .5)
x = matrix(x, nrow = 10) # 10 X 100 matrix
xs = apply(x, 2, scale)

x %*% t(x)
xs %*% t(xs)

v = t(xs) %*% xs / (10 - 1)
cor(xs[, 1:2])
```

## Least Squared Estimation

$$\hat{\beta} = (X^TX)^{-1}X^Ty$$

$$var(\hat{\beta}) = var((X^TX)^{-1}X^Ty) = var(y) (X^TX)^{-1}$$

## Standardized effect size

$$ Z = \frac{X^T Y}{\sqrt{N}}$$
```{r}
y = matrix(rnorm(1e3, 0, 1), nrow = 1e3)
x = matrix(rbinom(1e4, 2, .5), nrow = 1e3)
xs = apply(x, 2, scale)

p1 = t(xs) %*% y / sqrt(1e3)

p2 = apply(xs, 2, function(x1) {
  f1 = summary(lm(y ~ x1))
  f1$coefficients["x1", "Estimate"] / f1$coefficients["x1", "Std. Error"]
})

solve(t(xs) %*% xs) %*% t(xs) %*% y

plot(p1, p2); abline(0, 1)
```

## Genomic inflation factor 

y ~ $\lambda_{GC}$ * x, where x is the true $\chi^2$, y is the estimated $\chi^2$ of the association model

$\lambda_{GC}$ can be approximated as $median(y) / 0.456$

```{r}
S <- read.table(input,header=F)

if (stat_type == "Z") z = S[,1]
if (stat_type == "CHISQ") z = sqrt(S[,1])
if (stat_type == "PVAL") z = qnorm(S[,1]/2)

lambda = round(median(z^2)/.456,3)
```

## Meta-analysis using fixed effect model

```{r}
```

## Meta-analysis using random effect model

```{r}
```

## Conditional analysis using GWAS summary statistics and LD of a reference population

```{r}
```

## Imputation analysis using GWAS summary statistics and LD of a reference population

```{r}
```

## Transcriptome-wide association studies: association between expression of each gene with a trait

```{r}
```

## Burden test

$T_{burden} = w^T z$, where $w$ is a weight vector, $z$ is the effect size vector. 
Null distribution of $T_{burden}$ follows: $N(0, w^TVw)$.

```{r}
p = seq(1e-2, .99, 1e-2)
y = 1 / sqrt(p * (1-p))
plot(p, y)
```

## Overdispersion test: weighted sums of squared single-variant test statistics

```{r}
z = matrix(c(1:3), nrow = 3)
w = c(0.1, 2, 3) * diag(3)
t(z) %*% w %*% z
1 * .1 + 4 * 2 + 9 * 3
```

