## Variance components in regression models

? how to quantify variance explained by a predictor?
? and its relationship with estimated effect / standard error

## Variance definition

$$var(y) = E[y^2] - E[y]^2$$

## Law of total expectation

$$E[x] = E[E[x|y]]$$

## Variance definition under conditional distribution framework

$$E[y^2] - E[y]^2 = E[var(y|x)] + var(E[y|x]) $$

## Homoschedastic 

$var(y|x)$ does not depend on x, and is constant $\sigma^2$ across different values of x.

$$E[var(y|x)] = E[\sigma^2] = \sigma^2$$
$$var(y) = var(E[y|x]) + \sigma^2$$

## R function var() compute sample variance, which is an estimate of the population variance

Statistics uses random samples and test statistics to estimate population parameters, beause full population data is usually not available.
The sample average is usually a good estimate for the population average, but turns out the sample variance is a biased estimate for the population variance.
Sample variance tends to underestimate the population variance.
To estimate population variance in practice, given a sample vector $s$, we use: $\frac{n}{n-1} * (E[s^2] - E[s]^2)$.
However, this is an approximation, instead of a perfect correction.

```{r}
library(dplyr)
library(ggplot2)

myvar = function(x) { mean(x^2) - mean(x)^2 }

x = sample(c(0:1), size = 1e3, replace = T)
y = x + rnorm(1e3, 0, 1)

data = data.frame(x, y)
ggplot(data, aes(x = as.factor(x), y = y)) + geom_violin() + geom_boxplot(width = 0.1) + coord_flip()

myvar(y)

myvar(y[x == 0])
myvar(y[x == 1])

(sigma = mean(myvar(y[x == 0]), myvar(y[x == 1])))

mean(y[x == 0])
mean(y[x == 1])

(evar = myvar(c(mean(y[x == 0]), mean(y[x == 1]))))

evar = rep(mean(y[x == 0]), times = 1e3)
evar[x == 1] = mean(y[x == 1])

myvar(evar)

f = lm(y ~ x - 1)
summary(f)
sigma(f)

var(y)
var(predict(f))
var(unique(predict(f)))

var(predict(f)) + sigma(f)
```

## var(E[y|x]), explained variance, effect size, allele frequency

$$\beta^2 * var(x)$$

```{r}
x = sample(c(0:1), size = 1e3, replace = T)
y = 2 * x + rnorm(1e3, 0, 1)

f = lm(y ~ x)
summary(f)
beta = coefficients(f)["x"]

beta^2 * var(x)
var(predict(f))

var(y)
sigma(f)
sigma(f) + beta^2 * var(x)
sigma(f) + var(predict(f))

x1 = rbinom(1e3, 2, .3)
var(x1)
p = mean(x1) / 2
2 * p * (1-p)
```
